Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-22 doc/mc-paper.tex
CRAT --> CPOG

Computing certain properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as d-DNNF. \emph{Knowledge compilation} is the process of converting a formula
into such a form. Existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate the resulting model counts.

We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
of the representations used by existing knowledge compilers.
We have designed CRAT, a proof framework that can prove the equivalence of a
POG representation with a Boolean formula in conjunctive normal form (CNF),
the standard input format for SAT solvers and model counting tools.

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CRAT proofs certifying that the output POGs
are equivalent to the input CNF formulas. We have implemented an
efficient checker for these proofs. Our toolchain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CRAT
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

  %% CRAT
  \draw[thick] (40,12) [-latex] -- (50,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
  \node [above] at (44,12) {\texttt{.crat}};

@@

\end{tikzpicture}
}
\caption{Certifying toolchain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT)
  which can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CRAT file.
The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CRAT proof system, as well as correctness of its
concrete implementation, we formally verified the proof system as well as versions
of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CRAT file gives assurance that the proof
and the model count are correct. Our experience with developing a formally verified
proof checker has shown that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
In fact, subtle details in the partitioned sum rule were identified during this process.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
By contrast, our proofs construct a new propositional formula, and we must verify both that it
is partitioned and that it is equivalent to the input formula.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\theta$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\theta$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\theta$. We write (subscripted) $\theta$ for
clausal encoding formulas possibly using extension variables, and (subscripted) $\phi$ for formulas
that do not.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
from $\theta \cup \{C\}$ to create $\theta$.

@@

Compared to the proofs of unsatisfiability generated by SAT solvers,
ours have important differences.  Most
significantly, each proof step must preserve the set of solutions with respect to the input variables.
By contrast, an unsatisfiability proof need only guarantee that
no proof step cause a satisfiable set of clauses to become
unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.

\section{The CRAT POG Representation and Proof System}
\label{sect:crat}

%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

A CRAT file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
%  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
%  Generalizing to $k$ arguments would require
%  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \toprule
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
    \midrule
    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\
    \midrule
     & {\tt r} & $L$             &            & Declare root literal\\
    \bottomrule
  \end{tabular}
  }
\end{table}

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
%% The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.

@@

%% A CRAT proof follows the same general form as a QRAT dual
%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
%% addition and each clause deletion preserves equivalence.  With CRAT,
%% however, clauses are defined both explicitly and implicitly.  Starting
%% with the set of input clauses, the proof consists of a sequence of
%% steps that both add and delete clauses.  Each addition must be truth
%% preserving, that is, any satisfying total assignment for the set of clauses
%% before clause addition should still be a satisfying assignment afterwards.
%% Each deletion must be falsehood preserving.  That is, there can be no
%% new satisfying assignments when the clause is deleted.

@@

Appendix~\ref{app:crat:example} shows the CRAT description for
an input formula with five clauses, yielding a POG with five
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

\section{Generating CRAT from d-DNNF}
\label{section:generating:crat}

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
A d-DNNF
graph can be directly translated  into a POG,
although in doing this conversion,
our program also performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root node $\noder$ requires showing that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.
%%  Suppose at some point in this process that we want to
%% delete clause $C$ such that the remaining set of clauses will form the
%% CNF formula $\phi$.  Deleting $C$ requires proving that
%% $\phi \imply C$.

@@

{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The
earlier formulation of $\validate$ considers each literal $\lit \in L$
separately, calling a SAT solver for every literal that cannot be
validated with BCP\@.  Literal grouping handles
all of these literals together.  It defines product node $\nodev$
having the literals as children.  The goal then becomes to prove that
any total assignment must yield 1 for extension variable $v$.  Calling
a solver with $v$ set to 0 yields an unsatisfiability proof that can
be mapped back to a sequence of clause additions in the CRAT file validating
all of the literals.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
benchmark problems are available at \url{https://github.com/rebryant/crat}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our {\em verified}
versions of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our {\em prototype} versions are written in C
(checker) and Python (ring evaluator).  These are currently faster and
more scalable, but we anticipate their need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

Our proof generator is written in C/C++ and uses
\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
proof steps back into hinted CRAT clause additions, the generator can
use either its own RUP proof generator, or it can invoke
\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
to large proofs, but each invocation has a high startup cost.  We
therefore only use it when solving larger problems (currently ones
with over 1000 clauses).

@@

Finally, the runtimes for the MICE toolchain versus our CRAT toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general,
our CRAT toolchain showed better scaling, in part due to its ability
to control the recursion through lemmas.

@@

\section{CRAT Example}
\label{app:crat:example}

@@

\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\

>>>>

Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as a deterministic,
  decomposable, negation normal form (d-DNNF) graph.
\emph{Knowledge compilation} is the process of converting a formula
into such a form. Existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate the model counts.

We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
of the representations used by existing knowledge compilers.
We have designed  CPOG, a framework that can express proofs of equivalence between a
POG  and a Boolean formula in conjunctive normal form (CNF).

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CPOG (for ``checkable'' POG) file format. A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CPOG file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

  %% CPOG
  \draw[thick] (40,12) [-latex] -- (50,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
  \node [above] at (44,12) {\texttt{.cpog}};

@@

\end{tikzpicture}
}
\caption{Certifying toolchain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
  which can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CPOG file.
The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CPOG proof system, as well as correctness of its
concrete implementation, we formally verified the proof system as well as versions
of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CPOG file gives assurance that the proof
and the model count are correct. Our experience with developing a formally verified
proof checker has shown that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
In fact, subtle details in the partitioned sum rule were identified during this process.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
By contrast, our proofs construct a new propositional formula, and we must verify both that it
is partitioned and that it is equivalent to the input formula.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.

@@

We write (possibly subscripted) $\theta$ for
formulas encoded as clauses, possibly with extension variables.
We write (possibly subscripted)
$\phi$ for formulas
that use no extension variables.

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\theta$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\theta$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\theta$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:cpog}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
from $\theta \cup \{C\}$ to create $\theta$.

@@

Compared to the proofs of unsatisfiability generated by SAT solvers,
ours have important differences.  Most
significantly, each proof step must preserve the set of solutions with respect to the input variables.
By contrast, an unsatisfiability proof need only guarantee that
no proof step causes a satisfiable set of clauses to become
unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.

\section{The CPOG POG Representation and Proof System}
\label{sect:cpog}

%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

A CPOG file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
%  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
%  Generalizing to $k$ arguments would require
%  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

\begin{table}
  \caption{CPOG Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:cpog:syntax}
\centering{
  \begin{tabular}{lllll}
    \toprule
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
    \midrule
    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\
    \midrule
     & {\tt r} & $L$             &            & Declare root literal\\
    \bottomrule
  \end{tabular}
  }
\end{table}

Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
%% The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.

@@

%% A CPOG proof follows the same general form as a QRAT dual
%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
%% addition and each clause deletion preserves equivalence.  With CPOG,
%% however, clauses are defined both explicitly and implicitly.  Starting
%% with the set of input clauses, the proof consists of a sequence of
%% steps that both add and delete clauses.  Each addition must be truth
%% preserving, that is, any satisfying total assignment for the set of clauses
%% before clause addition should still be a satisfying assignment afterwards.
%% Each deletion must be falsehood preserving.  That is, there can be no
%% new satisfying assignments when the clause is deleted.

@@

Appendix~\ref{app:cpog:example} shows the CPOG description for
an input formula with five clauses, yielding a POG with five
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

\section{Generating CPOG from d-DNNF}
\label{section:generating:cpog}

%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
A d-DNNF
graph can be directly translated  into a POG,
although in doing this conversion,
our program also performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root node $\noder$ requires showing that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
CPOG framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.
%%  Suppose at some point in this process that we want to
%% delete clause $C$ such that the remaining set of clauses will form the
%% CNF formula $\phi$.  Deleting $C$ requires proving that
%% $\phi \imply C$.

@@

{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The
earlier formulation of $\validate$ considers each literal $\lit \in L$
separately, calling a SAT solver for every literal that cannot be
validated with BCP\@.  Literal grouping handles
all of these literals together.  It defines product node $\nodev$
having the literals as children.  The goal then becomes to prove that
any total assignment must yield 1 for extension variable $v$.  Calling
a solver with $v$ set to 0 yields an unsatisfiability proof that can
be mapped back to a sequence of clause additions in the CPOG file validating
all of the literals.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
benchmark problems are available at \url{https://github.com/rebryant/cpog}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our {\em verified}
versions of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our {\em prototype} versions are written in C
(checker) and Python (ring evaluator).  These are currently faster and
more scalable, but we anticipate their need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

Our proof generator is written in C/C++ and uses
\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
proof steps back into hinted CPOG clause additions, the generator can
use either its own RUP proof generator, or it can invoke
\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
to large proofs, but each invocation has a high startup cost.  We
therefore only use it when solving larger problems (currently ones
with over 1000 clauses).

@@

Finally, the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general,
our CPOG toolchain showed better scaling, in part due to its ability
to control the recursion through lemmas.

@@

\section{CPOG Example}
\label{app:cpog:example}

@@

\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
